Analysing Constraint Grammars with a SAT-solver

### 2016-03-01 paper/lrec/eval_futurework.tex
more text in intro & add label for conclusions
add another new feature; polish text a bit, actual content still TBD

\hline
                   & \textsc{nld}  & \textsc{spa}lem  & \textsc{spa}nolem \\ \hline
\# rules           & 59            & 279       & 279     \\ \hline
\# readings        & 336           & 3905      & 1735    \\ \hline
\# conflicts       & 7             & n         & m    \\ \hline
\clock{} all rules & ??s           & + 1h      & 30m-1h   \\ \hline
\clock{} last rule & ??s           & ??s       & ?min ??s    \\ \hline
\clock{} all rules 
without ambiguity 
classes            & 10ish sec       & ??min ??s    & ?h ??min    \\ \hline

@@

\paragraph{Spanish} The Spanish grammar had proportionately the highest number of errors.\footnote{TODO://link/to/error.log} \todo{confirm; I assume this is gonna be so, but let's see Fin and Por}.  The grammar we ran is like the one found in the Apertium repository, apart from 2 changes: we fixed some typos (capital O for 0) in order to make it compile, and commented out two rules that used regular expressions, because our solution does not handle them properly (see more on Future work/Implementation).

@@

The Spanish grammar has a high number of interaction conflicts, all with the same pattern:

@@

This is another instance of more general rule applying before more specific. Negated context works as well to make a rule stricter. The second rule will never apply, because the first will catch all cases.

In addition, there was a number of definitions that were never used. The program doesn't currently point them out to the user, but that is a trivial addition; just like rule-internal conflicts, it doesn't require any semantic analysis. 
Another error, which was not pointed by the program, but discovered when we tried to extract the readings from the grammar and use them

As with the Dutch grammar, we ran the tool on individual rules and examined the sequences that were generated. None of the following was marked as a conflict, but looking at the output indicated that

\begin{itemize}
\item[] 
\begin{verbatim}
REMOVE Vblex IF (0 "como") (*-1 Vblex BARRIER Cnj_Rel) ;
REMOVE Vblex IF (0 "como") (*1 Vblex BARRIER Cnj_Rel) ;
\end{verbatim}
\end{itemize}

Here is something that the ambiguity class constraint should find out.
The grammar writer wants to disambiguate the word form ``como'', which can be either a preposition, adverb or the first person singular of the verb ``comer''. 
However, the rule here addresses the lemma ``como'' instead of the word form, hence the SAT-solver cannot create a solution where \texttt{vblex} and ``como'' are in the same reading. The automated 0-to-target conversion could come in handy here too.

Side note: if this rule was written as \texttt{Vblex + "como"}, it would have been caught as a conflicting tag combination. ``Como'' as a lemma does not occur with \texttt{vblex}; it is only one word form which is ambiguous.

@@

For the Finnish grammar, we did not have time to investigate all the
results. We could verify approximately \nicefrac{1}{3} as true positives and
\nicefrac{1}{3} as false positives.
The verified true positives were all rule-internal conflicts. 
For example, the following rule requires the target to be genitive and unambiguously nominative: 
\begin{itemize}
\item[]\texttt{"oma" SELECT GEN IF [...] (0C NOM) ;}
\end{itemize}
We also detected a number of tagset errors;
the grammar was originally written in 1995, and converted to the
Omorfi tagset by \cite{pirinen2015}, but there were 15 rules that used the old convention for possessive suffixes.

%  For example, the Omorfi tagset
% marks a possessive suffix with \texttt{Px}, e.g. \texttt{PxPl2} for second person plural, but the
% grammar uses the old convention \texttt{PL2} in the rules.
% As a result, 15 rules with possessive suffixes were detected as
% impossible.
To improve performance, we ignored the fully specified readings from the
lexicon and used only tag combinations from the grammar:
for example, the rule \texttt{REMOVE (verb sg) IF (-1 det)}
gives us ``verb sg'' and ``det''.
The shortcut works most of the time, but it is possible to
miss some cases: e.g. \texttt{SELECT PRON + REL IF (0 NOM)} 
assumes ``pron rel nom'', but our method only gives
``pron rel'' and ``nom'' separately. 
17 false positives were caused by this shortcut. 
In addition, 2 false positives were caused by the naive implementation of
rules with \verb!*! in their context.

@@

\section{Future work}

>>>>

\hline
                   & \textsc{nld}  & \textsc{spa}lem  & \textsc{spa}nolem \\ \hline
\# rules           & 59            & 279       & 279     \\ \hline
\# readings        & 336           & 3905      & 1735    \\ \hline
\# conflicts       & 7             & n         & m    \\ \hline
\clock{} all rules & ??s           & + 1h      & 30m-1h   \\ \hline
% \clock{} last rule & ??s           & ??s       & ?min ??s    \\ \hline
\clock{} no amb. 
classes            & 10ish sec       & ??min ??s    & ?h ??min    \\ \hline

@@

\paragraph{Spanish} The Spanish grammar had proportionately the highest number of errors. \todo{confirm; I assume this is gonna be so, but let's see Fin and Por}.  The grammar we ran is like the one found in the Apertium repository, apart from 2 changes: we fixed some typos (capital O for 0) in order to make it compile, and commented out two rules that used regular expressions, because our solution does not handle them properly (see more on Future work/Implementation).

@@

The Spanish grammar has a high number of interaction conflicts, \todo{n} with the same pattern:

@@

This is another instance of more general rule applying before more
specific. Negated context works as well to make a rule stricter. The
second rule will never apply, because the first will catch all
cases. For a full list of found conflicts, see \todo{\url{actual://link.here/spanish/error.log}}

In addition, there was a number of definitions that were never
used. The program doesn't currently point them out to the user, but
that is a trivial addition; just like rule-internal conflicts, it does
not require any semantic analysis. 
Another error, which was not pointed by the program, but discovered when we tried to extract the readings from the grammar and use them

As with the Dutch grammar, we ran the tool on individual rules and
examined the sequences that were generated. None of the following was
marked as a conflict, but looking at the output indicated that there
are multiple interpretations, such as whether two analyses for a
context word should be in the same reading or different readings.
We observed also cases where the grammar writer has specified desired
behaviour in comments, but the rule does not do what the grammar
writer intended.

% \begin{itemize}
% \item[] 
% \begin{verbatim}
% REMOVE Vblex IF (0 "como") (*-1 Vblex BARRIER Cnj_Rel) ;
% \end{verbatim}
% \end{itemize}

% Based on the comments in the file, the grammar writer wants to
% disambiguate the word form ``como'', which can be either a
% preposition, adverb or the first person singular of the verb
% ``comer''. 
% However, the rule here addresses the \emph{lemma} ``como'' instead of the
% word form, hence the SAT-solver cannot create a solution where
% \texttt{vblex} and ``como'' as a lemma are in the same
% reading. 
% A precise definition for this rule would target \texttt{vblex "<como>"} 
% or \texttt{vblex "comer"}, so that it becomes obvious that those two
% are supposed to be in the same reading.

% Currently, as all the word forms are allowed to combine with anything,
% the ambiguity class constraint does not point this
% The automated 0-to-target conversion could come in handy here too.
% However, if the

% TODO An example of the same but in condition. that the ambiguity class constraint, when
% implemented properly, should find out.

@@

The Finnish grammar was the largest and most complex of all the
grammars we have tested.

the grammar was originally written in 1995, and converted to the
Omorfi tagset by \cite{pirinen2015}, but there were 15 rules that used the old convention for possessive suffixes.

To improve performance, we ignored the fully specified readings from the
lexicon and used only tag combinations from the grammar:
for example, the rule \texttt{REMOVE (verb sg) IF (-1 det)}
gives us ``verb sg'' and ``det''.
The shortcut works most of the time, but it is possible to
miss some cases: e.g. \texttt{SELECT PRON + REL IF (0 NOM)} 
assumes ``pron rel nom'', but our method only gives
``pron rel'' and ``nom'' separately. 
17 false positives were caused by this shortcut. 
In addition, 2 false positives were caused by the naive implementation of
rules with \verb!*! in their context.

@@

\section{Future work and conclusions}
\label{sec:conclusion}

The evaluation indicates that the tool finds cases that are not
obvious to the human eye.

One additional feature is to suggest reformattings for a rule. Recall
figure~\ref{fig:regroup} from the introduction; in that case, the
original rule was written by the original author, and another
grammarian thought that the latter form is nicer to read. Doing the
reverse operation could also be possible. If a rule with long
disjunctions conflicts, it may be useful to split it into smaller
conditions, and eliminate one at a time, in order to find the
reason(s) for the conflict.
